---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The repository containing zip data file was cloned from github.
```{r}
# unzipping the activity.zip file 

if(!file.exists('activity.csv')){
  unzip('activity.zip')
  file.remove('activity.zip')
}

```
```{r results='hide'}
library(dplyr)
library(ggplot2)
```

After checking the contents, date column was converted to date class.
```{r}
activity_raw <- read.csv('activity.csv',header = TRUE,na.strings = 'NA')

head(activity_raw)
str(activity_raw)
dim(activity_raw)
summary(activity_raw)

#converting date column to date data type
class(activity_raw$date)
activity_raw$date <- as.Date(activity_raw$date, format = '%Y-%m-%d')

```


## What is mean total number of steps taken per day?
finding distinct dates in the dataset.
```{r}
# finding number of distinct dates
length(unique(activity_raw$date))

```

Plotting the histogram and calculating the average number of steps 
```{r}
steps_per_day <- activity_raw %>% group_by(date) %>% 
  summarise(steps_per_day = sum(steps))

hist(steps_per_day$steps_per_day, main = 'Histogram of steps taken each day',
     xlab = 'steps per day', ylim = c(0,30), breaks = 6 )

```
```{r}
mean_steps <- mean(steps_per_day$steps_per_day, na.rm = TRUE)
median_steps <- median(steps_per_day$steps_per_day,  na.rm = TRUE)

mean_steps
median_steps
```
mean of the total number of steps taken per day is `r mean_steps`
median of the total number of steps taken per day is `r median_steps`

## What is the average daily activity pattern?

Grouping the data by inteval.
```{r}
average <- activity_raw %>% group_by(interval) %>% 
  summarise(average_steps = mean(steps, na.rm = TRUE))

average

```


```{r}
# time series plot 
ggplot(data = average, aes(x = interval, y = average_steps)) + geom_line() +
  ylab('Average steps') + labs(title = 'average steps per interval')
```

```{r}
#Which 5-minute interval, on average across all the days in the dataset, contains 
#the maximum number of steps?

maxINTERVAL <- average[which.max(average$average_steps),]$interval

maxINTERVAL
```


## Imputing missing values
```{r}
#Calculate and report the total number of missing values in the dataset
# (i.e. the total number of rows with NAs)

NA_data <- activity_raw[is.na(activity_raw$steps),] 
number_of_NA <- length(NA_data$steps)

number_of_NA

```

total number of missing rows is `r number_of_NA`

```{r}
# The strategy for filling NA

new_data <- activity_raw

for (i in 1:nrow(new_data)){
  if(is.na(new_data[i,'steps'])){
    new_data[i,'steps'] <- average[which(average$interval==new_data[i,'interval']),'average_steps']
    
  }
}

head(new_data)

```
 
plotting the histogram
```{r}
steps_wo_na <- new_data %>% group_by(date) %>% 
  summarise(sum_steps = sum(steps))

hist(steps_wo_na$sum_steps, main = 'Histogram of steps taken each day after filling NAs',
     xlab = 'steps per day', ylim = c(0,18), breaks = seq(0,23000,1000) )

```


```{r}
mean_steps_new <- mean(steps_wo_na$sum_steps, na.rm = TRUE)
median_steps_new <- median(steps_wo_na$sum_steps, na.rm = TRUE)

mean_steps_new
median_steps_new
```
* The new mean of the total number of steps taken per day is `r mean_steps_new`
* The new median of the total number of steps taken per day is `r median_steps_new`

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
The mean of number of steps has not increased. But there is a slight increase in median which is expected.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#creating weekday fun
weekday_type <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday")
  else 
    return("weekend")
}

new_data$date <- as.Date(new_data$date)

# creating factor variable
new_data$factor <- sapply(new_data$date, weekday_type)
```

subsetting by the weekday factor
```{r}
#calculating average steps per day for weekday and weekend
data_weekday <- new_data %>% filter(factor=='weekday') 
data_weekend <- new_data %>% filter(factor=='weekend') 


weekday_avg <- data_weekday %>% group_by(interval) %>% summarise(average = mean(steps))
weekend_avg <- data_weekend %>% group_by(interval) %>% summarise(average = mean(steps))

```

```{r}
# plotting time series
par(mfrow=c(2,1), mar=c(4,4,2,1))

plot(average~interval, data=weekday_avg, type="l",
       main="Avg steps for Weekdays", xlab="Time Interval", col="red")
plot(average~interval, data=weekend_avg, type="l",
       main="Avg steps for Weekends", xlab="Time Interval", col="red")

```



